Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-09-21 Tex/Paper.tex
Changes after discussion with Patric

% ******************************************************************************************************************** %
\section[\label{S:I}]{Introduction}
% interested from SMI/AWAKE
% requirements for AWAKE Run 2
% ******************************************************************************************************************** %

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic charged particle beam
travelling through a plasma excites a strong longitudinal electric field that can be loaded by a trailing witness beam.
With optimal beam loading, an increase in energy spread can be kept to a minimum. Acceleration of an electron withness
beam by an electron drive beam has been demonstrated experimentally in the past
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept proton driven plasma wakefield
accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the
accelerated beam. The beam will also see a transversely and longitudinally varying accelerating field causing an
increase in energy spread \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces an axially symmetric focusing force which preserves emittance \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam in the international linear collider with
$1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses as drivers, a large number of plasma
stages is required. Staging plasma accelerators without reducing the effective gradient and spoiling the beam quality is
challenging \cite{steinke:2016,lindstrom:2016}. Proton beams available at CERN carry significantly more energy,
$19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for driving longer plasma wakefield accelerator
stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications.
This issue is resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of
the buckets in the modulated structure. This self-modulation process is caused by the transverse fields generated by the
beam acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation
frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis with a
surrounding halo of defocused particles. The train resonantly produces wakefields to larger amplitudes.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

% ******************************************************************************************************************** %
\subsection[\label{S:I:AWAKE}]{AWAKE Run 2}
% ******************************************************************************************************************** %

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver and a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the transverse size
of the SPS proton beam such that $k_{pe}\sigma_{x} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number.
The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and energy spread.

@@

@@

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations \cite{berglyd_olsen:2015, berglyd_olsen:2016}.

@@

% ******************************************************************************************************************** %
\subsection[\label{S:M:Setup}]{Drive Beam Parameters}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}). When the
initial SPS beam containing $3\nexp{11}$ protons \cite{gschwendtner:2016} enters the second plasma cell, the peak
$E_{z}$ field is expected to be $500-600\funit{MV}{m}$. As the baseline AWAKE drive beam current is insufficient to
reach the non-linear regime and produce a plasma bubble, the plasma electrons are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions
are replicated in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

@@

% ******************************************************************************************************************** %
\subsection[\label{S:M:Setup}]{Witness Beam Parameters}
% ******************************************************************************************************************** %

Beam loading of a short witness beam is sensitive to its position relative to the field \cite{tzoufras:2009}. To
eliminate the problem of initial de-phasing of the witness beam with respect to the proton wakefield, initial beam
energy was set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. The relative drift causing the de-phasing is proportional
to $\gamma^{-2}$. By simply matching the initial $\gamma$ values we eliminate this effect entirely. However, a lower
initial value is likely to be sufficient \cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial normalised emittance of $\epsilon = 2\unit{\mu m}$,
corresponding to a $\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2\gamma}{\epsilon} = \frac{\lambda_{pe}}{2\pi}\sqrt{2\gamma}, \label{EQ:Matched}
\end{align}
This relation requires a very narrow beam compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$. The charge density
of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal beam
loading at around $1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one. We used a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal
grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles. For the largest parameter scans,
shown in figure \ref{Fig:BeamQ}, we used a transverse grid cell size of $2.34\unit{\mu m}$.

@@

% ******************************************************************************************************************** %
\section[\label{S:BL}]{Beam Loading}
%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]
%~
%~ Patric's Comments
%~ - Beam loading for narrow energy spread
%~ - Beam loading in linear regime (old papers from SLAC in the 80's, then Katsouleas for PWFA) and nonlinear regime
%~   (Tzoufras) well know
%~ - Propose "new regime" in the SM case, beam loading and blowout to get emittance preservation
%~ - Limitation as for PWFA driver, must use some of the W bunch to reach the blowout, it is kind of like the early SLAC
%~   experiments (84GeV) where the W-bunch is like the D-bunch in these experiments, but in a plasma prepared by the p+
%~   bunch and the SM
%~ - maybe also something about the fact that 200pC in a 60µm bunch corresponds to a 1kA current, much larger than the
%~   p+ bunch current, because of the fact that the wakefields are driven my multiple bunches?
% ******************************************************************************************************************** %

@@

While the ideal shape is trapezoidal, for Gaussian beams a reasonable flat field, and consequently low energy spread,
can be achieved by controlling the charge density to achieve a similar effect. This, however, puts the front of the
witness beam outside of the ideal region while the middle bulk of the beam has a trapezoidal-like shape. In our case,
with a matched beam to the plasma density defined by equation \ref{EQ:Matched}, we are limited on how much total charge
we can accelerate without reaching charge densities that will overload the field.

@@

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius within the bubble, where normalised emittance is preserved, follows the evolution given by equation
\ref{EQ:Matched} -- as shown in figure \ref{Fig:BeamFilament}. This has the potential to cause overloading of the field.
However, for our base case this effect was not significant.

@@

% ******************************************************************************************************************** %
\section[\label{S:C}]{Conclusion}
% ******************************************************************************************************************** %

@@

\section[\label{Ack}]{Acknowledgements}
% ******************************************************************************************************************** %

>>>>

% ******************************************************************************************************************** %
\section{Introduction}\label{S:I}
% interested from SMI/AWAKE
% requirements for AWAKE Run 2
% ******************************************************************************************************************** %

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic charged particle beam
travelling through a plasma excites a strong longitudinal electric field that can be loaded by a trailing witness beam.
With optimal beam loading, an increase in absolute energy spread can be kept to a minimum. Acceleration of an electron
withness beam by an electron drive beam has been demonstrated experimentally in the past
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept proton driven plasma wakefield
accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the
accelerated beam. The beam will also see a transversely and longitudinally varying accelerating field causing an
increase in energy spread \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces a focusing force that preserves emittance \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam that could be produced by the international
linear collider with $1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses as drivers, a
large number of plasma stages is required. Staging plasma accelerators without reducing the effective gradient and
spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}. Proton beams available at CERN carry
significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for driving longer plasma
wakefield accelerator stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for
such applications. This issue is resolved by letting the proton beam undergo self-modulation before injecting the
witness beam into one of the buckets in the modulated structure. This self-modulation process is caused by the
transverse fields generated by the beam acting upon the beam itself, causing regions of the beam to rapidly defocus
\cite{kumar:2010}. The modulation frequency is close to that of the plasma, and produces a train of short proton bunches
along the beam axis with a surrounding halo of defocused particles. The train resonantly produces wakefields to larger
amplitudes.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

% ******************************************************************************************************************** %
\subsection{AWAKE Run 2}\label{S:I:AWAKE}
% ******************************************************************************************************************** %

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver and a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the transverse size
of the SPS proton beam such that $k_{pe}\sigma_{x,y} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number.
The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and energy spread.

@@

[Replace]

@@

Our previous approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations \cite{berglyd_olsen:2015, berglyd_olsen:2016}.

@@

% ******************************************************************************************************************** %
\subsection{Drive Beam Parameters}\label{S:M:Setup}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}). When the
initial SPS beam containing $3\nexp{11}$ protons \cite{gschwendtner:2016} enters the second plasma cell, the peak
electric

field is expected to be $500-600\funit{MV}{m}$. As the baseline AWAKE drive beam current is insufficient to
reach the non-linear regime and produce a plasma bubble, the plasma electrons are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions
are replicated in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

@@

% ******************************************************************************************************************** %
\subsection{Witness Beam Parameters}\label{S:M:Setup}
% ******************************************************************************************************************** %

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}.
To eliminate the problem of initial de-phasing of the witness beam with respect to the proton wakefield, initial beam
energy was set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. The relative drift causing the de-phasing is proportional
to $\gamma^{-2}$. By matching the initial $\gamma$ values in these simulations we eliminate this effect entirely.
However, a lower initial value is likely to be sufficient \cite{berglyd_olsen:2015}.

In these simulations we have used a matched beam for an initial normalised emittance of $\epsilon_{N} = 2\unit{\mu m}$.
The beam matching relation is given by
\begin{equation}
    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
\end{equation}
where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
\begin{equation}
    \sigma_{x,y}^{2} = c\epsilon_{N}\sqrt{2\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}} \label{EQ:MatchedR}
\end{equation}
in order to be matched to a given plasma density, initial emittance and energy. This relation requires a very narrow
beam with a $\sigma_{x,y}$ of $5.25\unit{\mu m}$ compared to the drive beam $\sigma_{x,y} = 200\unit{\mu m}$. The charge
density of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal
beam loading at around $1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one. We used a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal
grid cells  for the simulations presented in section \ref{S:BL}. The witness beam was simulated with $16.8\unit{M}$
non-weighted particles. For the largest parameter scans, shown in figure \ref{Fig:BeamQ}, we used a transverse grid cell
size of $2.34\unit{\mu m}$.

@@

% ******************************************************************************************************************** %
\section{Beam Loading}\label{S:BL}
%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]
%~
%~ Patric's Comments
%~ - Beam loading for narrow energy spread
%~ - Beam loading in linear regime (old papers from SLAC in the 80's, then Katsouleas for PWFA) and nonlinear regime
%~   (Tzoufras) well know
%~ - Propose "new regime" in the SM case, beam loading and blowout to get emittance preservation
%~ - Limitation as for PWFA driver, must use some of the W bunch to reach the blowout, it is kind of like the early SLAC
%~   experiments (84GeV) where the W-bunch is like the D-bunch in these experiments, but in a plasma prepared by the p+
%~   bunch and the SM
%~ - maybe also something about the fact that 200pC in a 60µm bunch corresponds to a 1kA current, much larger than the
%~   p+ bunch current, because of the fact that the wakefields are driven my multiple bunches?
% ******************************************************************************************************************** %

@@

While the ideal shape is trapezoidal, for Gaussian beams a reasonable flat field, and consequently low energy spread,
can be achieved by controlling the charge density to achieve a similar effect. This, however, puts the front of the
witness beam outside of the ideal region while the middle bulk of the beam has a trapezoidal-like shape. In our case,
with a matched beam to the plasma density defined by equation \ref{EQ:MatchedB}, we are limited on how much total charge
we can accelerate without reaching charge densities that will overload the field.

@@

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
transverse beam size within the bubble, where normalised emittance is preserved, follows the evolution given by equation
\ref{EQ:MatchedR} where the plasma density is given by the ion density -- as shown in figure \ref{Fig:BeamFilament}.
This has the potential to cause overloading of the field. However, for our base case this effect was not significant.

@@

% ******************************************************************************************************************** %
\section{Conclusion}\label{S:C}
% ******************************************************************************************************************** %

@@

\section{Acknowledgements}\label{Ack}
% ******************************************************************************************************************** %
